---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
head(df_titanic)
```

**Observations**:

- The variables are class, sex, age, child, survived (survival status), and n (number).

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
  - There were 2208 souls aboard the Titanic.
- Are there any differences?
  - The dataset details 2201 people. There are seven people missing from the dataset.
  - Relative to the total number of people aboard, 7 individuals are relatively negligible (\<1%), however, depending on the status of the 7 missing individuals, the data may be greatly affected (ie. if there were 2 children aboard, and the excluded data points were all children, the statistical analysis may change significantly).
- If yes, what might account for those differences?
  - It is possible that there was not data on these individuals, so they could not be accounted for.
  - It is possible that the individuals did not fit into the data set somehow, and thus were excluded.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(Class, n, fill = Sex)) +
  geom_col(position = "stack")
```

**Observations**:

- More Crew members survived than 1st Class members.
- In the 1st and 2nd classes more women survived than men. This is by strict count, I would be curious to know a.) what proportion of each class was male/female, and b.) what proportion of each sex survived by class for further context.
- More 3rd class members survived than second class. Again, I wonder about the relative populations for a more apples-to-apples comparison.
- Of the classes, Crew had the fewest females survive (by count). I am curious about the initial sex distribution among the crew.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "stack") +
  
  labs(y = "Proportion Survived (%)") +
  scale_y_continuous(labels = scales::label_percent())
```

**Observations**:

- Write your observations here.
  - The proportion of survived women to survived men is more equal in the first and second classes, and more dominant in the 3rd class and crew.
- Is there anything *fishy* going on in your plot?
  - The proportions amount to greater than 1 (100%) of a category.
  - There is a warning about removed rows containing "missing values or values outside of the scale range."

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(Class, Prop, fill = Sex, alpha = Age)) +
  geom_col(position = "dodge") +
  
  scale_alpha_manual(values = c("Adult" = 1, "Child" = 0.4)) +
  labs(y = "Proportion Survived (%)") +
  scale_y_continuous(labels = scales::label_percent())
```

**Observations**:

- The children in 1st and 2nd class had a 100% survival rate according to the graph.
- There were no children on the crew.
- The survival rate of female children in first class is lower than that of female adults in first class.
- Female adult passengers had a much higher survival rate (proportions greater than 2x that of adult male passengers in the same class).
- Adult males in 2nd class had the lowest proportion of survivors.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - I suspect the the q4 plot had proportions greater than 100% because it treated children and adults as different categories with independent proportions.
  - There is still the missing values warning.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
